### 2018-06-20 manuscript.rmd
added a sentence
added additional sup fig

```{r preliminaries}
library( tidyverse )
library( gridExtra )
library( ape )
library( hutan )
library( magrittr )
library( doParallel )
library( png )

@@

@@

@@

ppred =
	lapply(
		ppred_files,
		function( ppred_file ){
			row = str_match( ppred_file, "WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.ppred" )  # "\\((.*?) :: (0\\.[0-9
			dim(row) = NULL
			lines = read_lines( ppred_file )
			if ( ! all( lines[ expected_test_name_lines ] == expected_test_names ) ){
				stop("ppred file not of expected structure")
			}
			z_scores =
				lines[ expected_zscore_lines ] %>%
				str_replace( "z-score : ", "") %>%
				as.numeric()

			D = data.frame(
				file = row[ 1 ],
				transform = row[ 2 ],
				states = as.numeric( row[ 3 ] ),

@@

				# name in ppred file: diversity test
				# phylobaesmpi flag: -div, "performs a posterior predictive diversity test: the test statistic is the mean diversity per site"
				# Feuda et al name: ppa-div, "site-specific amino acid diversity... measuring whether models can adequately estimate... the mean number of distinct amino acids observed at each site"
				ppa_div = z_scores[ 1 ],

@@

			return( D )
		}
	) %>%
	bind_rows()

D =
	ppred %>%
	gather( ppred, score, ppa_div:ppa_mean)

#D %>%
#	ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
#		geom_line()+
#		geom_point()+
#		scale_x_reverse() +
#		facet_grid( . ~ ppred )

D$transform = paste(rep("random-", length(D$transform)), D$transform, sep="")
D$transform[D$transform == "random-none"] = "SR6-reproduced"

@@

D = bind_rows( D, original )

D$transform = as.factor(D$transform)

#use more descriptive PP names
D$ppred[D$ppred=="ppa_div"] = "PP diversity"
D$ppred[D$ppred=="ppa_conv"] = "PP convergence"
D$ppred[D$ppred=="ppa_var"] = "PP site hetero"
D$ppred[D$ppred=="ppa_max"] = "PP taxon hetero max"
D$ppred[D$ppred=="ppa_mean"] = "PP taxon hetero mean"

@@

	ctenosis = get_clade_frequency(trees, clade_ctenosis)
	porisis = get_clade_frequency(trees, clade_porisis)
	D = data.frame(
		file = row[ 1 ],
		transform = row[ 2 ],
		states = as.numeric( row[ 3 ] ),
		ctenosis = ctenosis,
		porisis = porisis
	)

	return( D )
}

@@

tree_support$transform = factor( tree_support$transform, levels=levels(D$transform) )
```

```{r plots_pp}
plot_ppa = function( ppred_focal, D, show_legend=FALSE ){
	g_plot = 
		D %>%
			filter( ppred==ppred_focal ) %>% 
			ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
			geom_line()+
			geom_point()+
			scale_x_reverse(breaks=breaks) +
			ggtitle(ppred_focal) +
			expand_limits(y=c(0, NA)) +
			theme_classic()

@@

# g_ppa = sapply( unique(D$ppred), function(ppred){ plot_ppa( ppred, D ) } )

ppreds = unique(D$ppred)

#grid.arrange(
#	plot_ppa( ppreds[1], D, all_legends),
#	plot_ppa( ppreds[2], D, all_legends),
#	plot_ppa( ppreds[3], D, all_legends),
#	plot_ppa( ppreds[4], D, all_legends),
#	plot_ppa( ppreds[5], D, all_legends),
#	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
#	nrow = 2,
#	ncol = 3
#)

@@

D %<>% filter( states %in% states_to_consider)
tree_support %<>% filter( states %in% states_to_consider)

@@

g = arrangeGrob(
	g_legend( plot_ppa( ppreds[5], D, TRUE ) ),
	plot_tree_support,
	plot_ppa( ppreds[3], D, all_legends),
	plot_ppa( ppreds[2], D, all_legends),
	plot_ppa( ppreds[1], D, all_legends),
	plot_ppa( ppreds[4], D, all_legends),
	plot_ppa( ppreds[5], D, all_legends),
	layout_matrix = lay
)

# grid.arrange( g )

@@

In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A). Feuda *et al.* [1] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis. Their analyses consider datasets from studies by Chang *et al.* [2] and Whelan *et al.* [3], both of which found support for Ctenophora-sister. Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies. In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids. These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [4]. The intent is to discard many lineage-specific changes, which are expected to fall within these groups. They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies". Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide significant support for Porifera-sister. Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister. The new analyses show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.

@@

This conclusion does not follow from their own presented results. Only a single analysis with PP scores provides what could be considered strong support (>95% posterior probability) for Porifera-sister. Of the 32 analyses, PP scores were calculated for 16 (those for the full Whelan and Chan matrices). Based on PP scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses (Supplemental Figure 1). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which PP scores are available provides significant support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix. Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 1). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 1), indicating that it did not converge as well as other analyses.

@@

The statistics presented in Supplemental Figure 1 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display in R.

### Supplemental figure

![](figures/Supplemental_Figure_1_manuscript.png)

**Supplemental Figure 1.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded according to whether support for Ctenophora-sister (red) or Porifera-sister (green) is greater than 95% posterior probability, or unresolved (the posterior probabilities for neither hypothesis are greater than 95%). Only one of these analyses, the SR6 coding of the Whelan matrix, has >95% support for Porifera-sister. The 20-state and 6-state points on the plots in Figure 1 of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five PP statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides significant support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.

>>>>

```{r preliminaries}
library( tidyverse )
library( gridExtra )
library( ape )
library( hutan )
library( magrittr )
library( doParallel )
library( png )
library( scales )

@@

posterior_prob_thresh = 0.95

@@

```{r load_table3_data}

D = read.table( "tidy_table_3.tsv", header=TRUE, stringsAsFactors=FALSE  )

D$Model = factor( D$Model, c("GTR+G", "CAT-GTR+G" ) )

D$Dataset = sub( "D20", "", D$Dataset )
D$Dataset = sub( "Chang", "Chang-Holo", D$Dataset )
D$Dataset = sub( "Opistho", "Opis", D$Dataset )
D$Dataset = sub( "Choano", "Cho", D$Dataset )
D$Dataset = factor( D$Dataset, c( "Whelan-Cho", "Whelan-Holo", "Whelan-Opis", "Chang-Holo" ) )

D$Recoding[ D$Recoding == "None" ] = "20"
D$Recoding[ D$Recoding == "KGB-6" ] = "KGB6"
D$Recoding[ D$Recoding == "Dayhoff-6" ] = "D6"
D$Recoding[ D$Recoding == "SR-6" ] = "SR6"
D$Recoding = factor( D$Recoding, c("20", "KGB6", "D6", "SR6" ) )

D$Result = "Unresolved"
D$Result[ D$Porifera_sister >= posterior_prob_thresh ] = "Porifera-sister"
D$Result[ D$Ctenophora_sister >= posterior_prob_thresh ] = "Ctenophora-sister"
D$Result = factor( D$Result )

D$Convergence = "bad"
D$Convergence[ D$Maxdiff <= 1 ] = "possible"
D$Convergence[ D$Maxdiff <= 0.3 ] = "acceptable"
D$Convergence[ D$Maxdiff <= 0.1 ] = "good"
D$Convergence = factor( D$Convergence, c( "good", "acceptable", "possible", "bad" ) )

D$Maxdiff = round( D$Maxdiff, 2 )

cteno_color = "#F8766D"
poro_color = "#00BA38"
unresolved_color = "#6F94CD"

```{r result_support}

gg_support = 
D %>%
	ggplot( aes( Recoding, Dataset ) ) +
		geom_tile(aes(fill = Porifera_sister)) +
		facet_wrap(~ Model) +
		scale_fill_gradient(low = cteno_color, high = poro_color, limits=c(0, 1)) +
		geom_text(aes(label = Porifera_sister)) +
		ggtitle("Porifera-sister Support") +
		theme_classic()

ggsave( 
	file="figures/Supplemental_Figure_1_manuscript_raw.pdf", 
	device="pdf", 
	width = 6.85, 
	height = 4, 
	dpi = 300, 
	gg_support )

```{r result_sub}

D_sub =
	D %>%
	rename( Porifera_sister_posterior = Porifera_sister ) %>%
	filter(Model=="CAT-GTR+G") %>%
	filter( (Dataset=="Whelan-Opis") | (Dataset=="Chang-Holo") ) %>%
	select (-c(Ctenophora_sister, Burnin, effsize, rel_diff)) %>%
	gather( var, val, Porifera_sister_posterior:PPA_MEAN ) %>%
	mutate( 
		var = 
			factor( 
				var, 
				levels = c(
					"Porifera_sister_posterior", 
					"Cycles", 
					"Maxdiff", 
					"PPA_DIV", 
					"PPA_MAX", 
					"PPA_CONV", 
					"PPA_VAR", 
					"PPA_MEAN")
				)
		)

gg_sub =
	D_sub %>%
		ggplot( aes( Recoding, Dataset ) ) +
			geom_tile( aes( fill = Result ) ) +
			facet_wrap( ~ var, ncol = 2 ) +
			geom_text(aes(label = val), size=3) +
			scale_fill_manual( values=c( cteno_color, poro_color,unresolved_color ) ) +
			ggtitle("Variation between analyses") +
			theme_classic()

ggsave( 
	"figures/Supplemental_Figure_2_manuscript_raw.pdf", 
	gg_sub, 
	device=pdf, 
	width=6.5, 
	height=6 
	)

@@

ppred =
	lapply(
		ppred_files,
		function( ppred_file ){
			row = 
				str_match( 
					ppred_file, 
					"WhelanD20_AA_mod_transform-([0-9a-zA-Z]+)_states-([0-9]+).chain.ppred" 
					)  # "\\((.*?) :: (0\\.[0-9

			dim(row) = NULL
			lines = read_lines( ppred_file )
			if ( ! all( lines[ expected_test_name_lines ] == expected_test_names ) ){
				stop("ppred file not of expected structure")
			}
			z_scores =
				lines[ expected_zscore_lines ] %>%
				str_replace( "z-score : ", "") %>%
				as.numeric()

			Dpp = data.frame(
				file = row[ 1 ],
				transform = row[ 2 ],
				states = as.numeric( row[ 3 ] ),

@@

				# name in ppred file: diversity test
				# phylobaesmpi flag: -div, "performs a posterior predictive diversity test: 
				#   the test statistic is the mean diversity per site"
				# Feuda et al name: ppa-div, "site-specific amino acid diversity... 
				#  measuring whether models can adequately estimate... the mean number of 
				#  distinct amino acids observed at each site"
				ppa_div = z_scores[ 1 ],

@@

			return( Dpp )
		}
	) %>%
	bind_rows()

Dpp =
	ppred %>%
	gather( ppred, score, ppa_div:ppa_mean)

#Dpp %>%
#	ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
#		geom_line()+
#		geom_point()+
#		scale_x_reverse() +
#		facet_grid( . ~ ppred )

Dpp$transform = paste(rep("random-", length(Dpp$transform)), Dpp$transform, sep="")
Dpp$transform[Dpp$transform == "random-none"] = "SR6-reproduced"

@@

Dpp = bind_rows( Dpp, original )

Dpp$transform = as.factor(Dpp$transform)

#use more descriptive PP names
Dpp$ppred[Dpp$ppred=="ppa_div"] = "PP diversity"
Dpp$ppred[Dpp$ppred=="ppa_conv"] = "PP convergence"
Dpp$ppred[Dpp$ppred=="ppa_var"] = "PP site hetero"
Dpp$ppred[Dpp$ppred=="ppa_max"] = "PP taxon hetero max"
Dpp$ppred[Dpp$ppred=="ppa_mean"] = "PP taxon hetero mean"

@@

	ctenosis = get_clade_frequency(trees, clade_ctenosis)
	porisis = get_clade_frequency(trees, clade_porisis)
	Dts = data.frame(
		file = row[ 1 ],
		transform = row[ 2 ],
		states = as.numeric( row[ 3 ] ),
		ctenosis = ctenosis,
		porisis = porisis
	)

	return( Dts )
}

@@

tree_support$transform = factor( tree_support$transform, levels=levels(Dpp$transform) )
```

```{r plots_pp}
plot_ppa = function( ppred_focal, Dpp, show_legend=FALSE ){
	g_plot = 
		Dpp %>%
			filter( ppred==ppred_focal ) %>% 
			ggplot( aes(x=states, y=score, group=transform, col=transform) ) +
			geom_line()+
			geom_point()+
			scale_x_reverse(breaks=breaks) +
			ggtitle(ppred_focal) +
			expand_limits(y=c(0, NA)) +
			theme_classic()

@@

# g_ppa = sapply( unique(Dpp$ppred), function(ppred){ plot_ppa( ppred, Dpp ) } )

ppreds = unique(Dpp$ppred)

#grid.arrange(
#	plot_ppa( ppreds[1], Dpp, all_legends),
#	plot_ppa( ppreds[2], Dpp, all_legends),
#	plot_ppa( ppreds[3], Dpp, all_legends),
#	plot_ppa( ppreds[4], Dpp, all_legends),
#	plot_ppa( ppreds[5], Dpp, all_legends),
#	g_legend( plot_ppa( ppreds[5], Dpp, TRUE ) ),
#	nrow = 2,
#	ncol = 3
#)

@@

Dpp %<>% filter( states %in% states_to_consider)
tree_support %<>% filter( states %in% states_to_consider)

@@

g = arrangeGrob(
	g_legend( plot_ppa( ppreds[5], Dpp, TRUE ) ),
	plot_tree_support,
	plot_ppa( ppreds[3], Dpp, all_legends),
	plot_ppa( ppreds[2], Dpp, all_legends),
	plot_ppa( ppreds[1], Dpp, all_legends),
	plot_ppa( ppreds[4], Dpp, all_legends),
	plot_ppa( ppreds[5], Dpp, all_legends),
	layout_matrix = lay
)

# grid.arrange( g ) # To render the plot directly in the text

@@

In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Figure 1A). Feuda *et al.* [1] present new phylogenetic analyses that they claim provide conclusive support for the Porifera-sister hypothesis. Their analyses consider datasets from studies by Chang *et al.* [2] and Whelan *et al.* [3], both of which found support for Ctenophora-sister. Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies. In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids. These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [4]. The intent is to discard many lineage-specific changes, which are expected to fall within these groups. Rather than model compositional heterogeneity, this approach discards heterogeneous information so that much simpler models with fewer states can be applied. They report that posterior predictive (PP) analyses [5] indicate recoded analyses have better model adequacy than amino acid analyses, and "Porifera-sister was favored under all recoding strategies". Here I focus on two aspects of Feuda *et al.* First, I point out that many of their recoded analyses are actually unresolved, and that the analyses with the best PP scores do not provide strong support for Porifera-sister. Second, I present new analyses that show that the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition across lineages. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister. The new analyses show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.

@@

This conclusion does not follow from their own presented results. Only a single analysis with PP scores provides what could be considered strong support (&ge; `r posterior_prob_thresh` posterior probability) for Porifera-sister. Of the 32 analyses, PP scores were calculated for 16 (those for the full Whelan and Chan matrices). Based on PP scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses (Supplemental Figure 2). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which PP scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [4] recoded Whelan [3] matrix. Furthermore, this analysis does not have the best score according to any of the five PP statistics they considered (Supplemental Figure 2). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Supplemental Figure 2), indicating that it did not converge as well as other analyses. It should also be noted that none of the reanalyses of Chang matrix provide strong contradiction of that original study. Chang *et al.* [2] found strong support for Ctenophora-sister, and all of the reanalyses of this same matrix by Feuda *et al.* also have strong support for Ctenophora-sister or are unresolved.

@@

The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code on the source file for this document.

### Supplemental figures

```{r result_sig, fig.width=6.5, fig.height=3.1}

gg_sig =
	D %>%
		ggplot( aes( Recoding, Dataset ) ) +
			geom_tile( aes( fill = Result ) ) +
			facet_wrap( ~ Model ) +
			theme_classic()

#ggsave( "figures/figure_3.pdf", gg_sig, device=pdf, width=6.5, height=3.1 )

gg_sig

```
**Supplemental Figure 1.**  A grapchical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3. Cells are color coded by whether posterior probability is &ge; `r posterior_prob_thresh` for Porifera-sister,  &ge; `r posterior_prob_thresh` for Ctenophora-sister, or neither (unresolved). Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure.

![](figures/Supplemental_Figure_2_manuscript.png)

**Supplemental Figure 2.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Figure 1. Only one of these analyses, the SR6 coding of the Whelan matrix, has &ge; `r posterior_prob_thresh` support for Porifera-sister. The 20-state and 6-state points on the plots in Figure 1 of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five PP statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the PP scores, and showed the poorest convergence according to Maxdiff.
